---
title: "IS607: Data Acquisition and Management Final Project"
author: "Jason Joseph"
date: "December 1, 2015"
output:
  html_document:
    toc: true
    theme: spacelab
    highlight: textmate
---

```{r echo=FALSE, message=FALSE,warning=FALSE}
setwd("C:\\GradSchool\\607 Proj")
library(XML)
library(sqldf)
library(ggplot2)
library(stringr)
library(dplyr)
library(tidyr)
```



# Introduction

The education sector is one of the most important areas of development in any country. The educational system is the foundation on which a country's human resource is moulded so they can make a contribution to the country's development. Year after year governments touts on the increased amount of money allocated and spent on their educational system. The government of Guyana also touts on increased spending in the education sector, but still the majority of Guyana's population remains uneducated. I am curious to discover whether the amount spent on education influences a country's literacy rate, pupil to teacher ratio and the number of students enroled in primary, secondary and tertiary schools.

# Data Acquisition

## Functions

### World Bank API

The World Bank database was my primary source where I retrieved my data. So I found it fitting to write a function to retrieve, clean and transform the data from the World Bank website via its API.  

```{r}
fetchIndicatorValues <- function(indicator)
{
  url.base <- sprintf("http://api.worldbank.org/countries/all/indicators/%s?format=xml&per_page=10000",indicator)
  doc.xml <- xmlParse(url.base)
  
  # get root element of the data set
  doc.xml.root <- xmlRoot(doc.xml)
  
  ## getting number of pages from the xml root attribute
  pages <- as.integer(xmlAttrs(doc.xml.root)[["pages"]])
  
  # creating an empty dataframe to store the query results
  data.set <- data.frame(indicator = character(0),country = character(0),value = double(0),decimal=integer(0),stringsAsFactors = FALSE)
  
  # iterating over the pages and storing the query restults into the data.set dataframe
  for(i in 1:pages)
  {
    url.page <- sprintf("%s&page=%s",url.base,i) 
    doc.xml <- xmlParse(url.page)
    doc.xml.root <- xmlRoot(doc.xml)
    doc.xml.elements <-xmlChildren(doc.xml.root)
    doc.df <- xmlToDataFrame(doc.xml.elements,stringsAsFactors = FALSE)
    data.set <- rbind(data.set,doc.df)
  }
  
  # Transforming the value column from chracter to double class type
  data.set$value <- as.double(data.set$value)
  # Transforming the date column from chracter to integer class type
  data.set$date <- as.integer(data.set$date)
  #removing unwanted columns
  data.set$indicator <- data.set$decimal <- NULL
  # removing records with NA values
  data.set <- na.omit(data.set)
  # Removing continent and special country grouping
  data.set <- subset(data.set, !(country %in% group.items$country))
  # removing rownames so a recount of the row numbers can be triggered
  row.names(data.set) <- NULL
  # Replacing ' from country names to an empty space 
  data.set$country <- str_replace_all(data.set$country,"'"," ")
  return(data.set)
} 

```

### Calculating Mean Values

The datasets contain values for each country for several years. For this project and the analysis that follows later down in the pipeline. I needed to calculate the mean value and expenditure over that time frame. The function below does exactly that. 

```{r}

calculateMeanVals <- function(countryName,tableName, col1,col2,col3 = NA)
{
 if(is.na(col3))
    {
     col3 <- ""
    }
    else
    {
     col3 <- sprintf(",%s",col3)
    }
  
  # Select the records for a country and calculates the min/max date and average value and expenditure for that period
  
  df <- sqldf(sprintf("Select  country,min(date) as minDate,max(date) as maxDate ,AVG(%s)as %s,AVG(%s)as %s 
                      %s
                      FROM 
                      '%s' where country = '%s'
                      GROUP BY country %s"
                      ,col1, 
                      sprintf("avg_%s",col1) ,
                      col2,
                      sprintf("avg_%s",col2),
                      col3,
                      tableName,
                      countryName,
                      col3))
  
  return(df)
}
```
### List of grouped regions

The datasets contained values for grouped regions e.g. Caribbean small states, Euro area, etc. I created a text file that contains the grouped regions that will be used to filter out those values from the datasets.

```{r}
# list of regions that are grouped in the dataset
group.items <- read.table("https://raw.githubusercontent.com/jasonjgy2000/IS607/master/Projects/607%20Proj/Data/expList.txt",stringsAsFactors = FALSE,header = FALSE,sep = "\t",col.names = c("country"))

```

## Government expenditure on education, total (% of government expenditure)

Retrieving the general government expenditure on education (current, capital, and transfers) is expressed as a percentage of total general government expenditure on all sectors and storing it into a data frame.

```{r}
gov.exp.df <- fetchIndicatorValues("SE.XPD.TOTL.GB.ZS")
head(gov.exp.df)
```


## Literacy rate, adult total (% of people ages 15 and above)

Retrieving and cleaning the adults (15+) literacy rate percentage dataset. It contains the total percentage of the population age 15 and above who can, with understanding, read and write a short, simple statement on their everyday life. 


```{r}

literacy.rate.df <- fetchIndicatorValues("SE.ADT.LITR.ZS")
head(literacy.rate.df)

literacy.rate.expenditure <-dplyr::inner_join(gov.exp.df,literacy.rate.df,by=c("country","date"))

colnames(literacy.rate.expenditure) <- c("country","date","expenditure","literacyrate")


countries <- literacy.rate.expenditure[!duplicated(literacy.rate.expenditure[,"country"]),1]

  for(i in 1: length(countries))
  {
    if(i == 1)
    {
      literacy.rate.expenditure.avg <-  calculateMeanVals(countries[i],"literacy.rate.expenditure","expenditure","literacyrate")
    }
    else
    {
literacy.rate.expenditure.avg <- rbind(literacy.rate.expenditure.avg,calculateMeanVals(countries[i],"literacy.rate.expenditure","expenditure","literacyrate"))  
    }
    
  }  

head(literacy.rate.expenditure.avg)
#summary(literacy.rate.expenditure.avg)
```


## Pupil to teacher ratio in primary Schools

Below I am parsing the number of pupils enrolled in primary school divided by the number of primary school teachers. This dataset was parsed from a CSV file, cleaned and transformed into a data frame.


```{r}
# load Csv
pupil.teacher.df <- read.csv(file="https://raw.githubusercontent.com/jasonjgy2000/IS607/master/Projects/607%20Proj/Data/se.prm.enrl.tc.zs_Indicator_en_csv_v2.csv",skip = 4,header = TRUE, stringsAsFactors = FALSE,check.names = FALSE)

# removing columns not needed 
pupil.teacher.df$`Country Code` <- pupil.teacher.df$`Indicator Name` <- pupil.teacher.df$`Indicator Code` <-  NULL

# renaming first column from  Country name to country
colnames(pupil.teacher.df)[1] <- "country"

#removing extra column with NA's
pupil.teacher.df[,58] <- NULL

dplyr::tbl_df(pupil.teacher.df)

# wide to long transformation
pupil.teacher.df <- pupil.teacher.df %>%  gather("date","value",2:57)

# change Year column from factor to integer class type
pupil.teacher.df$date <- as.integer(as.character(pupil.teacher.df$date))

# removing records with NA values
pupil.teacher.df <- na.omit(pupil.teacher.df)

# Removing continent and special country grouping
pupil.teacher.df <- subset(pupil.teacher.df, !(country %in% group.items$country))

# removing rownames so a recount of the row numbers can be triggered
row.names(pupil.teacher.df) <- NULL

# Replacing ' from country names to an empty space 
pupil.teacher.df$country <- str_replace_all(pupil.teacher.df$country,"'"," ")

# joining the pupil teacher ratio data frame to the expenditure data frame by country and date
pupil.techer.expenditure <-dplyr::inner_join(gov.exp.df,pupil.teacher.df,by=c("country","date"))

# assigning new column names to the pupil teacher ratio expenditure data frame
colnames(pupil.techer.expenditure) <- c("country","date","expenditure","pupilTeacher")

# selecting distinct country names from the pupil.teacher.expenditure data frame
countries <- pupil.techer.expenditure[!duplicated(pupil.techer.expenditure[,"country"]),1]

# creating the average pupil teacher expenditure data frame.

  for(i in 1: length(countries))
  {
    if(i == 1)
    {
      pupil.techer.expenditure.avg <-  calculateMeanVals(countries[i],"pupil.techer.expenditure","expenditure","pupilTeacher",NA)
    }
    else
    {
pupil.techer.expenditure.avg <- rbind(pupil.techer.expenditure.avg,calculateMeanVals(countries[i],"pupil.techer.expenditure","expenditure","pupilTeacher",NA))  
    }
  }  


head(pupil.techer.expenditure.avg)
```

## Student enrollment primary, secondary and tertiary 

Retrieving and cleaning the datasets that contain the gross enrolment ratio for primary, secondary and tertiary schools.

```{r warning=FALSE,message=FALSE}

enrollment.primary <- fetchIndicatorValues("SE.PRM.ENRR")
head(enrollment.primary)

enrollment.secondary <- fetchIndicatorValues("SE.SEC.ENRR")
head(enrollment.secondary)

enrollment.tertiary <- fetchIndicatorValues("SE.TER.ENRR")
head(enrollment.tertiary)

enrollment.primary$type <- "primary"
enrollment.secondary$type <- "secondary"
enrollment.tertiary$type <- "tertiary"

# combining the primary, secondary and tertiary data frames into one data frame

enrollment.all <- rbind(enrollment.primary,enrollment.secondary,enrollment.tertiary)

# joining the enrollment data frame to the expenditure data frame by country and date
enrollment.expenditure <-dplyr::inner_join(gov.exp.df,enrollment.all,by=c("country","date"))

colnames(enrollment.expenditure) <- c("country","date","expenditure","enrollment","type")

# selecting distinct country names from the enrollment.all data frame
countries <- enrollment.all[!duplicated(enrollment.all[,"country"]),1]

# creating the average entrollment expenditure data frame. 
  for(i in 1: length(countries))
  {
    if(i == 1)
    {
      enrollment.expenditure.avg <-  calculateMeanVals(countries[i],"enrollment.expenditure","expenditure","enrollment","type")
    }
    else
    {
enrollment.expenditure.avg <- rbind(enrollment.expenditure.avg,calculateMeanVals(countries[i],"enrollment.expenditure","expenditure","enrollment","type"))  
    }
    
  } 

head(enrollment.expenditure.avg)
```
#Visualization

## Government expenditure on education, total (% of government expenditure)

```{r}
ggplot(gov.exp.df, aes(x = value)) + geom_histogram() +ggtitle("Government Expenditure")
```

## Literacy rate, adult total (% of people ages 15 and above)
```{r}
plot(literacy.rate.expenditure.avg$avg_literacyrate~literacy.rate.expenditure.avg$avg_expenditure, xlab = "Expenditure",ylab = "Literacy Rate",main ="Government Expenditure vs Literacy Rate")
```

## Pupil to teacher ratio in primary Schools
```{r}
plot(pupil.techer.expenditure.avg$avg_pupilTeacher~pupil.techer.expenditure.avg$avg_expenditure, xlab = "Expenditure",ylab = "Pupil Teacher Ratio",main ="Government Expenditure vs Pupil Teacher Ratio")
```


## Student enrollment primary, secondary and tertiary 

```{r}
plot(enrollment.expenditure.avg$avg_enrollment~enrollment.expenditure.avg$avg_expenditure, xlab = "Expenditure",ylab = "Student Enrollment",main ="Government Expenditure vs Student Enrollment")
```


# Analysis

## Literacy rate, adult total (% of people ages 15 and above)
```{r}
literacy.rate.cor <- cor.test(literacy.rate.expenditure.avg$avg_expenditure,literacy.rate.expenditure.avg$avg_literacyrate)

literacy.rate.cor$estimate
```


## Pupil to teacher ratio in primary Schools

```{r}
pupil.teacher.cor <-  cor.test(pupil.techer.expenditure.avg$avg_expenditure,pupil.techer.expenditure.avg$avg_pupilTeacher)
pupil.teacher.cor$estimate
```

## Student enrollment primary, secondary and tertiary 

```{r}
student.enrollment.cor <-  cor.test(enrollment.expenditure.avg$avg_expenditure,enrollment.expenditure.avg$avg_enrollment)
student.enrollment.cor$estimate
```

```{r message=FALSE,warning=FALSE}

df <- data.frame(item = character(0), corvalue = double(0), stringsAsFactors = FALSE)
df <- rbind(df,data.frame(item = "Literacy Rate",value = literacy.rate.cor$estimate , stringsAsFactors = FALSE))
df <- rbind(df,data.frame(item = "Pupil To Teacher Ratio",value = pupil.teacher.cor$estimate , stringsAsFactors = FALSE))
df <- rbind(df,data.frame(item = "Student Enrollment",value = student.enrollment.cor$estimate , stringsAsFactors = FALSE))
row.names(df) <- NULL
ggplot(df, aes(x = item, y=value)) + geom_histogram(stat = "identity") + ggtitle("Correlation")
```

# Conclusion

There is an extremely weak relationship between government expenditure into the educational system and their country's literacy rate, pupil to teacher ratio and the number of students enrolled in primary, secondary and tertiary schools.

