---
title: "Regex Assignment - Week 4"
author: "Jason Joseph"
date: "September 20, 2015"
output: html_document
---

```{r echo=FALSE}
library(stringr)
```

1) Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression.

a. [0-9]+\\\\$

     This expression is looking for any number of digits preceding a backslash. With the backslash being the end of the string. e.g. **1234\\**
     
b. \\\\b[a-z]{1,4}\\\\b
    
    This expression is looking for a string with a backslash followed by a **"b"** then  one to four letters with another **"b"** following. The last **"b"** doesn't have to be the last element in the string. e.g. **\\boys\\being**
    

c. .*?\\.txt$ 

    This expression is validating a string with any alphanemeric chracters followed by a backslash then by any character then txt. With txt being the end of the string. e.g.**"hello\\stxt"**
    
    

d. \\\\d{2}/\\\\d{2}/\\\\d{4}

    The expression is validating a backslash followed by two d's preceding a foward slash followed by a backslash and two d's followed by another forwardslash and four d's. e.g.**"\\dd\/\\dd\/\\dddd"**
    
e.   <(.+?)>.+?<\/\\\\1>  

    This expression is looking for a string that begins with character/s surrounded by a less-than and more than sign preceding that can be any character/word/s which is followed by a less-than sign, a backslash then the number one followed by the more-than sign.  e.g. **"\<hey1234>test<\/\\1>"**


2) Rewrite the expression [0-9]+\\\\$ in a way that all elements are altered but the expression performs the same task. **"\^\\d*[\\]\$"**

  
```{r}
#  "^\\d*[\\]$"
pat <- "^\\d*[\\\\]"
st <- "1234\\"
str_match(st,pat)
```


3 Consider the mail address chunkylover53[at]aol[dot]com

a) Transforming the email address to a standard mail format.
```{r}
# step 1 find and replace [at] with @
r <- "chunkylover53[at]aol[dot]com"
pat1 <- "\\[at\\]"
r <- str_replace(r,pat1,"@")

r

# step2 find and replace [dot] with .
pat1 <- "\\[dot\\]"
r <- str_replace(r,pat1,".")
r
```

b) Using [:digit:] as the regular expression pattern would fail because it would extract the first number it encounters. To fix this we append "+"which will allow us to extract all the digits in the email address.

Before Correction
```{r}
pat1 <- "[:digit:]"
str_extract(r,pat1)
```
After Correction
```{r}
pat1 <- "[:digit:]+"
str_extract(r,pat1)
```

c) Using "\\\\D" as the regular expression to extract the numbers in the email address would fail because it would identify the first character in the string that is not a number.The correct expression would be "\\\\d+".

Before Correction
```{r}
pat1 <- "\\D"
str_extract(r,pat1)
```
After Correction
```{r}
pat1 <- "\\d+"
str_extract(r,pat1)
```


