---
title: "U.S. International Trade Data"
author: "Jason Joseph"
date: "October 7, 2015"
output: html_document
---
# Questions for  Analysis: 

1). Calculate the average goods export for each month over the three-year span.

2). Calculate the average goods import for each month over the three-year span.

3). Calculate the average services export for each month over the three-year span.

4). Calculate the average services import for each month over the three-year span.

```{r echo=FALSE}
library(dplyr)
library(tidyr)
```
# Import Data 

Thanks to this artice from stackoverflow, which gived me the idea of replacing missing values with NA.

http://stackoverflow.com/questions/13822801/whats-the-best-way-to-replace-missing-values-with-na-when-reading-in-a-csv

```{r}
data_set <- read.csv("https://raw.githubusercontent.com/jasonjgy2000/IS607/master/Projects/Project%202/International%20Trade%20in%20Goods%20and%20Services.csv",skip = 2,na.strings = c("","NA"))
head(data_set)
```

# Data Cleansing
```{r}
#assigning user-friendly column names
colnames(data_set) <- c("Period","GoodsType","2013-Export","2013-Import","2014-Export","2014-Import","2015-Export","2015-Import")
# removing empty rows
data_set <- filter(data_set,!is.na(data_set$GoodsType))
# fill missing data - Thanks Joy! 
name_rows<-seq(from = 1, to = nrow(data_set), by=2)
nameless_rows<-seq(from = 2, to = nrow(data_set), by=2)
data_set[nameless_rows,1]<-data_set[name_rows,1]
head(tbl_df(data_set))

```
# Data Transformation
```{R}
data_set<- data_set %>%  gather("Year","Amount",3:8) %>%  separate("Year",c("Year","Transaction"),sep="-")
# Remove "," from Amount figures. Thanks Youqing!
data_set$Amount <- as.numeric(gsub(',','',data_set$Amount))
head(tbl_df(data_set))
```

# Analysis

Question 1 : Calculate the average goods export for each month over the three-year span.

```{r}

mMean <- data_set %>% filter(!GoodsType == "Services" & Transaction =="Export") %>% group_by(Period) %>% summarise(pmean = mean(Amount))
mMean
```


Question 2 : Calculate the average goods import for each month over the three-year span.

```{r}

mMean <- data_set %>% filter(!GoodsType == "Services" & Transaction =="Import") %>% group_by(Period) %>% summarise(pmean = mean(Amount))
mMean
```

Question 3 : Calculate the average services export for each month over the three-year span.

```{r}

mMean <- data_set %>% filter(GoodsType == "Services" & Transaction =="Export") %>% group_by(Period) %>% summarise(pmean = mean(Amount))
mMean
```



Question 4 : Calculate the average services import for each month over the three-year span.

```{r}

mMean <- data_set %>% filter(GoodsType == "Services" & Transaction =="Import") %>% group_by(Period) %>% summarise(pmean = mean(Amount))
mMean
```