---
title: "Project 1"
author: "Jason Joseph"
date: "September 22, 2015"
output: html_document
---
```{r echo=FALSE}
library(RCurl)
library(stringr)
library(foreign)
```
 **Overview**

The objective of this project is to import and transform the structure of a text file with contains the results of a  chess tournament and produce a CSV output of the results. 
My game plan for this project was to import the results into a data frame. I would then massage the frame's structure to match the layout required for the  CSV output. To achieve the desired result the following steps were taken:

**Step 1**: Import Chess Tournament Result Data into R


```{r}
#Load chess tournament data 
data_set <- read.table("https://raw.githubusercontent.com/jasonjgy2000/IS607/master/Projects/Week%204%20Project/tournamentinfo.txt",skip = 3,sep = "|",fill = TRUE,header = FALSE)
head(data_set)
```

**Step 2**: Remove rows that were filled with the line separator "---" that was used in the text file. I then  removed the extra column that was created during the import.

```{r}
#Remove rows filled with seperating character "---"
data_set <- subset(data_set,!grepl("-{1,}",data_set$V1))
# Remove extra column (V11) and row names column
data_set$V11 <- data_set$row.names <- NULL
head(data_set)
```

**Step 3**: I assigned user-friendly column names to the data frame.

```{r}
# assign column names to data frame
colnames(data_set) <- c("Pair_num","Player_Name", "Total Number of Points","Round_1","Round_2","Round_3","Round_4","Round_5","Round_6","Round_7")
head(data_set)
```

**Step 4**:  I extracted each player's pre- rating score and merged them into the row
with the player's information under the column heading "pre-rating". I then removed the extra rows.

```{r}
# Extract players pre-rating into from seperate column and merge
exp <- "([R]\\:\\s*[[:digit:]]+)"
n <- 1
while(n < nrow(data_set))
{
  data_set$state[n] <- toString(data_set$Pair_num[n+1])
  data_set$pre_rating[n] <- str_extract(str_extract(data_set$Player_Name[n+1],exp),"[[:digit:]]+")
  data_set$state[n+1] <- NA
  data_set$pre_rating[n+1] <- NA
  n <- n+2
}

# remove extra rows 
data_set <- subset(data_set,grepl("[[:digit:]]",data_set$Pair_num))
head(data_set)
```

**Step 5**: I created a function that accepts each player's round data and returns the average pre-chess rating of their opponents that was assigned to the opp_aver column within the data frame.

```{r}
#calcuate mean 
exp <- "[[:digit:]]+"

calMean <- function(x1,x2,x3,x4,x5,x6,x7){
  items <- c(as.numeric(data_set$pre_rating[as.numeric(str_extract(x1,exp))]),
             as.numeric(data_set$pre_rating[as.numeric(str_extract(x2,exp))]),
             as.numeric(data_set$pre_rating[as.numeric(str_extract(x3,exp))]),
             as.numeric(data_set$pre_rating[as.numeric(str_extract(x4,exp))]),
             as.numeric(data_set$pre_rating[as.numeric(str_extract(x5,exp))]),
             as.numeric(data_set$pre_rating[as.numeric(str_extract(x6,exp))]),
             as.numeric(data_set$pre_rating[as.numeric(str_extract(x7,exp))]))
  
  return (round(mean(items,na.rm = TRUE)))
  
}

data_set$opp_aver <- mapply(calMean,data_set$Round_1,data_set$Round_2,data_set$Round_3,data_set$Round_4,data_set$Round_5,data_set$Round_6,data_set$Round_7)
head(data_set)
```
**Step 6**: All non-essential columns were removed. I then renamed each column to reflect those required in the CSV output. I then reorder the columns to match the ordering provided for the CSV.

```{r}
# remove extra columns 
data_set$row.names <- data_set$Pair_num <- data_set$Round_1 <- data_set$Round_2 <- data_set$Round_3 <- data_set$Round_4 <- data_set$Round_4 <- data_set$Round_5  <- data_set$Round_6 <- data_set$Round_7 <- NULL
row.names(data_set) <- NULL
# rename columns 
colnames(data_set) <- c("Player's Name","Total Number of Points","State","Player's Pre-Rating","Average Pre Chess Rating of Opponents")

# reorder columns 
data_set <- data_set[c(1,3,2,4,5)]
head(data_set)
```
**Step 7**: Now that the structure of the data frame matches the structure required for the CSV output, the "write.CSV" function was used to output the data frame into the CSV format. 

```{r}
# create csv
write.csv(data_set,"output.csv")
```